Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-16 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

Many datasets include individuals from multiple settings, geographic
regions, or even different studies. In the simplest case, individuals
(e.g., students) are nested within so-called clusters (e.g., school
classes). More complex clustered structures may occur when there are
multiple hierarchical levels (e.g., patients within hospitals within
regions or countries), or when the clustering is non-nested (e.g.,
electronic health record data from diverse settings and populations
within large databases). In general, individuals from the same cluster
tend to be more similar than individuals from other clusters. In
statistical terms, this implies that observations from the same cluster
are correlated. If this correlation is left unaddressed, estimates of
\emph{p} values, confidence intervals even model parameters are prone to
bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
require adequate handling and propagation of variance; we are not
recommending the adoption of multilevel models for data analysis here,
but rather for imputation. VMTdJ: ``I would only make that link after
introducing missing data, and missing data is not mentioned yet, so I
would do that in the next section''{]} Statistical methods for clustered
data typically adopt hierarchical models that explicitly describe the
grouping of observations. These models are also known as `multilevel
models', `hierarchical models', `mixed effect models', `random effect
models', and in the context of time-to-event data as `frailty models'.
Table \ref{tab:clus} provides an overview of some key concepts in
multilevel modeling.

@@

Like any other dataset, clustered datasets are prone to missing data.
Several strategies can be used to handle missing data, including
complete case analysis and imputation. We focus on the latter approach
and discuss statistical methods for replacing the missing data with one
or more plausible values. Afterwards, the completed data can be analyzed
as if they were completely observed. In contrast to single imputation
(where missing data are only replaced once), multiple imputation allows
to preserve uncertainty due to missingness and is therefore recommended
(c.f. Rubin 1976).

When clustered datasets are affected by missing values, we can
distinguish between two types of missing data: sporadic missingness and
systematic missingness \citep{resc13}. Sporadic missingness arises when
variables are missing for some but not all of the units in a cluster
\citep{buur18, jola18}. For example, it is possible that test results
are missing for several students in one or more classes. {[}TODO:
Provide an example for one of the case studies below.{]} When all
observations are missing within one or more clusters, data are
systematically missing. {[}TODO: Refer to Figure 1 and put
interpretation in the figure caption. VMTdJ: ``Mention in text that this
is sporadic missingness?''{]}

@@

Imputation of missing data requires consideration of the mechanism
behind the missingness. Rubin proposed to distinguish between data that
are missing completely at random (MCAR), data that are missing at random
(MAR) and data that are missing not at random (MNAR; see Table
\ref{tab:miss}). For each of these three missingness generating
mechanisms, different imputation strategies are warranted \citet{yuce08}
and \citet{hox15}. We here consider the general case that data are MAR,
and expand on certain MNAR situations.

@@

The \proglang{R} package \pkg{mice} has become the de-facto standard for
imputation by chained equations, which iteratively solves the
missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. However, commonly used imputation methods were not
designed for use in clustered data and usually generate observations
that are independent whereas multilevel data are dependent. For this
reason, we discuss how the \proglang{R} package \pkg{mice} can be used
to impute multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. We provide practical guidelines and
code snippets for different missing data situations, including
non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
imputation by chained equations with \pkg{mice} exclusively; other
imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
outside the scope of this tutorial. {[}TODO: change this sentence
because we do not assume familiarity: ``Assumed knowledge includes basic
familiarity with multilevel imputation \citep[see e.g.][ and
\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
models (see Table \ref{tab:mod})''{]}.

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(miceadds)     # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
R> library(miceheckman)  # for imputation
\end{CodeInput}
\end{CodeChunk}

>>>>

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. {[}TODO: make a link to imputation methods, which
require adequate handling and propagation of variance; we are not
recommending the adoption of multilevel models for data analysis here,
but rather for imputation. VMTdJ: ``I would only make that link after
introducing missing data, and missing data is not mentioned yet, so I
would do that in the next section''{]} Statistical methods for clustered
data typically adopt hierarchical models that explicitly describe the
grouping of observations. These models are also known as `multilevel
models', `hierarchical models', `mixed effect models', `random effect
models', and in the context of time-to-event data as `frailty models'.
Table \ref{tab:clus} provides an overview of some key concepts in
multilevel modeling.

@@

As with any other dataset, clustered datasets may be impacted by
missingness in much the same way. Several strategies can be used to
handle missing data, including complete case analysis and imputation. We
focus on the latter approach and discuss statistical methods for
replacing the missing data with one or more plausible values. Imputation
seperates the missing data problem from the analysis and the completed
data can be analyzed as if it were completely observed. It is generally
recommended to impute the missing values more than once to preserve
uncertainty due to missingness and to allow for valid inferences (c.f.
Rubin 1976).

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. {[}TODO: Provide an example for one of
the case studies below.{]} When all observations are missing within one
or more clusters, data are said to be systematically missing. {[}TODO:
Refer to Figure 1 and put interpretation in the figure caption. VMTdJ:
``Mention in text that this is sporadic missingness?''{]}

@@

Imputation of missing data requires consideration of the mechanism
behind the missingness. Rubin proposed to distinguish between data that
are missing completely at random (MCAR), data that are missing at random
(MAR) and data that are missing not at random (MNAR; see Table
\ref{tab:miss}). For each of these three missingness generating
mechanisms, different imputation strategies are warranted
(\citet{yuce08} and \citet{hox15}). We here consider the general case
that data are MAR, and expand on certain MNAR situations.

@@

{[}GV: DELETE THIS SECTION?{]}However, commonly used imputation methods
were not designed for use in clustered data and usually generate
observations that are independent whereas multilevel data are dependent.
For this reason, we discuss how the \proglang{R} package \pkg{mice} can
be used to impute multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
standard for imputation by chained equations, which iteratively solves
the missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}.

We provide practical guidelines and code snippets for different missing
data situations, including non-ignorable mechanisms. For reasons of
brevity, we focus on multilevel imputation by chained equations with
\pkg{mice} exclusively; other imputation methods and packages (e.g.,
\pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
{[}TODO: change this sentence because we do not assume familiarity:
``Assumed knowledge includes basic familiarity with multilevel
imputation \citep[see e.g.][ and \citet{grun18}]{audi18} and the
\pkg{lme4} notation for multilevel models (see Table
\ref{tab:mod})''{]}.

@@

\begin{itemize}
\tightlist
\item
  GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf heckman models
\end{CodeInput}
\end{CodeChunk}
